\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-20 paper.tex
Yans updates

Fig.~\ref{fig-ancestry-resolution}C therefore represents a \emph{sample-resolved} gARG,
% useful to say that we don't *need* to simulate an eARG and convert to a
% sample-resolved form, but can simulate a sample-resolved gARG directly
of the sort that can be simulated directly by the Hudson algorithm.
This differs in some important ways from its equivalent eARG (Fig.~\ref{fig-ancestry-resolution}A).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events. Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this resolution of ancestral material once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.

@@

The graph visualisations in Fig.~\ref{fig-simplification} have
some novel features, which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child).
% This is useful why? I can't come up with a quick way of saying
% this. In some senses it's a measure of identifiability?
% JK: Hmm, not sure we want to jump into inferability here?
All these features will impact what kind of an ARG we can
hope to infer from sampled genomes.

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} genetic sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods based on
(A) parsimony, (B) sampling the SMC$^\prime$, (C) copying from
inferred ancestral haplotypes, and (D) correlated local tree estimation.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
X and Y positions of graph nodes are arbitrary;
line width and node colour are as described in Fig.~\ref{fig-simplification}.
Graphics on the bottom row show the genomic positions, relative to the start of the ADH gene,
for each graph edge from the corresponding ARG. Edge intervals are drawn as horizontal lines,
stacked in time order (edges with youngest children at the bottom);
vertical dashed lines indicate breakpoints between local trees.
}
\end{figure}

@@

To illustrate the varying levels of information that can be captured in a gARG,
Fig.~\ref{fig-inferred-args} presents graph representations of genealogies output by two tools that
\emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and 
\argweaver~\citep{rasmussen2014genome}, 
and two which do not (\tsinfer\ and \relate). 
All inferences are based on the
same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
assume a mutation rate of 
% per site per generation may be easier to follow?
\SI{5.49e-9}{per site per generation} and a recombination rate of
\SI{8.4e-9}{per site per generation} \citep{lauterbur2022expanding}, together with
a constant effective population size of 1,000,000.
The version numbers of inference software, together with the specific inference parameters
needed to reproduce the ARGs, are available from our GitHub repository.

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args} A\& B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples 
from the posterior distribution over
ARGs given the data and a fixed model (the SMC$^\prime$, an approximation of the full coalescent
with recombination \citep{marjoram2006fast}).
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args} C \& D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args} A \& B, each recombination node has exactly 2 parents,
corresponding to a single recombination event with a specific breakpoint. Assuming that each
breakpoint occurs a a different location (the "infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with greater
than two parents, indicating deliberate imprecision: these are cases where multiple recombination events
have been ``pushed down'' onto a single more recent node. Note, however, that the number
of breakpoints in these two ARGs is smaller than the number of ARG nodes with multiple parents,
because those nodes frequently share breakpoint positions. The sharing of breakpoint locations
between multiple-parent nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.

Shading by coalescent span reveals that many non-recombinant (green) nodes are
``locally unary'' (see section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome, these nodes are present in the ancestry but do
not correspond to coalescence. As argued previously, this can provide potentially useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate\ ARG: other than the samples it has entirely solid nodes.
Only the \tsinfer\ results have nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the lack of precision in recombination nodes with more
than two parents. The clearest example is the polytomy involving the nodes
\noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}, which happen to
have identical sequences. \kwarg, \argweaver, and \relate\ require
bifurcating trees, so are forced
to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
relationship between these samples). 
The number of nodes in Fig.~\ref{fig-inferred-args} C
that have more than two children also hints at considerable uncertainty in
the order of branching elsewhere in the ARG.
% TODO add explanation of the bottom row.
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
%The example of successive simplification of a simulated ARG
%in Fig.~\ref{fig-simplification} illustrates the point
%that is is possible to \emph{store} arbitrary levels of precision
%about recombination in a gARG.
%It is also possible to \emph{infer} various levels of precision about
%recombination.
%To illustrate this, and to demonstrate the structural
%heterogeneity of the ARGs inferred  by current methods,
%Fig.~\ref{fig-inferred-args} shows the results of ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
%for the classical \citet{kreitman1983nucleotide} benchmark dataset
%using  \kwarg~\citep{ignatieva2021kwarg},
%\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
%\tsinfer~\citep{kelleher2019inferring},
%and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
%The minimum number
%of recombination breakpoints required in the absence of recurrent mutation
%is 7 for this dataset~\citep{song2003parsimonious},
%and the number of inferred breakpoints
%under the input mutation and recombination rates
%varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
%The ARGs inferred by the different methods also have major differences
%in how recombination information is represented.
%Essentially,
%\kwarg\ and \argweaver\ infer eARGs that contain
%specific recombination \emph{events} for each breakpoint.
%On the other hand, \tsinfer\ and
%\relate\ do not infer events but rather
%infer gARGs with by differing levels of precision about recombination.

The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
We can see, for example, that the \relate\ trees ``share'' less topology
than other methods, but it certainly does not infer a series of
\emph{unrelated} local trees
%(i.e., with either unlabelled internal
%nodes or differently labelled nodes in each tree).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees; for example,
both infer that Af-f, Fr-f and Wa-f share a common ancestor along
the entire sequence.
They contain a level of detail about recombination
that lies somewhere between a sequence of unrelated local trees on one
extreme and a completely precise event ARG on the other.

>>>>

Fig.~\ref{fig-ancestry-resolution}C therefore represents a \emph{sample-resolved} gARG,
% useful to say that we don't *need* to simulate an eARG and convert to a
% sample-resolved form, but can simulate a sample-resolved gARG directly
of the sort that can be simulated directly by the Hudson algorithm.
This differs in some important ways from its equivalent eARG (Fig.~\ref{fig-ancestry-resolution}A).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{q} is normally omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events. Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this resolution of ancestral material once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.

@@

The graph visualisations in Fig.~\ref{fig-simplification} have
some novel features, which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child). Nodes which
are never locally unary therefore have a coalescent span of 100\%,
whereas nodes in which ancestral material never coalesces have a
coalescent span of 0\% (such nodes will be extremely difficult to
identify from the distribution of shared mutations in the samples).
% This is useful why? I can't come up with a quick way of saying
% this. In some senses it's a measure of identifiability?
% JK: Hmm, not sure we want to jump into inferability here?
All these features will impact what kind of an ARG we can
hope to infer from sampled genomes.

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} genetic sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods based on
(A) parsimony, (B) sampling the SMC$^\prime$, (C) copying from
inferred ancestral haplotypes, and (D) correlated local tree estimation.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
X and Y positions of graph nodes are arbitrary;
line width and node colour are as described in Fig.~\ref{fig-simplification}.
Graphics on the bottom row show the genome positions, relative to the start of the ADH gene,
for each graph edge from the corresponding ARG. Edge intervals are drawn as horizontal lines,
stacked in time order (edges with youngest children at the bottom);
vertical dashed lines indicate breakpoints between local trees.
}
\end{figure}

@@

To illustrate the varying levels of information that can be captured in a gARG,
Fig.~\ref{fig-inferred-args} presents graphical representations of genealogies output by two tools that
\emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and 
\argweaver~\citep{rasmussen2014genome}, 
and two which do not (\tsinfer\ and \relate). 
All inferences are based on the
same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
assume a mutation rate of 
% per site per generation may be easier to follow?
\SI{5.49e-9}{per site per generation} and a recombination rate of
\SI{8.4e-9}{per site per generation} \citep{lauterbur2022expanding}, together with
a constant effective population size of 1,000,000.
To focus on the ARG topology rather than estimates of node times, the Y position of each node
is arbitrary, taken from the ranked order of nodes output by each software method.
The version numbers of inference software, the specific inference parameters
needed to reproduce the ARGs, and (where necessary) functions to convert outputs to the
\texttt{tskit} gARG format, are available from our GitHub repository.

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args} A\& B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples 
from the posterior distribution over
ARGs given the data and a fixed model (the SMC$^\prime$ \citep{marjoram2006fast},
an approximation of the full coalescent with recombination).
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args} C \& D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

% Here we describe recombination features
A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args} A \& B, each recombination node has exactly 2 parents,
corresponding to a single recombination event with a specific breakpoint. Assuming that each
breakpoint occurs a a different location (the "infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with greater
than two parents, indicating deliberate imprecision: these are cases where multiple recombination events
have been ``pushed down'' onto a single more recent node.
Note, however, that the number
of breakpoints in these two ARGs is smaller than the number of ARG nodes with multiple parents,
because those nodes frequently share breakpoint positions. The sharing of breakpoint locations
between multiple-parent nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.

% Here we move on to coalescence
As for coalescence in these inferred ARGs, the shading (``coalescent span'')
reveals that many non-recombinant (green) nodes are
locally unary (see section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome these nodes are present in the ancestry without
corresponding to coalescence of ancestral material. As argued previously, this can provide useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate\ ARG: other than the samples it has entirely solid nodes.
Only the \tsinfer\ results have nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the imprecision associated with recombination nodes having more
than two parents. The clearest example is the polytomy involving the nodes
\noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}, which happen to
have identical sequences. \kwarg, \argweaver, and \relate\ require
bifurcating trees, so are forced
to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
relationship between these samples). 
The number of nodes in Fig.~\ref{fig-inferred-args} C
that have more than two children also hints at considerable uncertainty in
the order of branching elsewhere in the ARG.

%% possible TODO - mention something about multiple local roots in
% tsinfer and Relate versus the other two.

%The example of successive simplification of a simulated ARG
%in Fig.~\ref{fig-simplification} illustrates the point
%that is is possible to \emph{store} arbitrary levels of precision
%about recombination in a gARG.
%It is also possible to \emph{infer} various levels of precision about
%recombination.
%To illustrate this, and to demonstrate the structural
%heterogeneity of the ARGs inferred  by current methods,
%Fig.~\ref{fig-inferred-args} shows the results of ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
%for the classical \citet{kreitman1983nucleotide} benchmark dataset
%using  \kwarg~\citep{ignatieva2021kwarg},
%\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
%\tsinfer~\citep{kelleher2019inferring},
%and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
%The minimum number
%of recombination breakpoints required in the absence of recurrent mutation
%is 7 for this dataset~\citep{song2003parsimonious},
%and the number of inferred breakpoints
%under the input mutation and recombination rates
%varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
%The ARGs inferred by the different methods also have major differences
%in how recombination information is represented.
%Essentially,
%\kwarg\ and \argweaver\ infer eARGs that contain
%specific recombination \emph{events} for each breakpoint.
%On the other hand, \tsinfer\ and
%\relate\ do not infer events but rather
%infer gARGs with by differing levels of precision about recombination.

% TODO add additional explanation of the bottom row, if enough isn't written below
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
We can see that both \tsinfer and (to a lesser extent) \relate have edges
that span multiple tree boundaries, indicating that they certainly
do not infer a series of \emph{unrelated} local trees.
%(i.e., with either unlabelled internal
%nodes or differently labelled nodes in each tree).
For example, both infer that \noderef{Af-f}, \noderef{Fr-f} and \noderef{Wa-f}
share a common ancestor along
the entire sequence. They do not, however, show extensive node sharing in the
oldest time periods, reflecting the fact that they do not encode topology
above the local tree roots (i.e. they do not go back to a
"grand MRCA"). This is also seen in the fact they their ARGs contain multiple roots
(nodes with zero parents, in blue), whereas \kwarg and \argweaver have a
single ARG root.
These two methods contain a level of detail about recombination
that lies somewhere between a sequence of unrelated local trees on one
extreme and a completely precise event ARG on the other.
